Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-12-01 paper/paper.tex
Up borrador VF

\setstretch{1.5}
This document was last modified at 2025-11-29 23:44:43 and it was last
rendered at 2025-11-29 23:44:43.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions embed market-based criteria of ``fair'' allocation and
shape how citizens understand deservingness and the balance between
state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for the allocation of essential services such as
healthcare, education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they legitimize unequal outcomes as the
product of individual responsibility and promote the conception of basic
social services as commodities.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}). As many countries expanded mandatory or voluntary
individual-capitalisation schemes---most radically in Chile---pensions
became organised around contribution-based entitlements and market
performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes not only how pension institutions allocate old-age risk and tie
benefit levels to lifetime labour-market trajectories
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also transforms the
justice principles citizens associate with old-age security: they must
judge whether higher benefits should follow market-derived criteria or
solidaristic principles
(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Although we know
something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: people with higher socioeconomic status are more
likely to support market-based criteria than disadvantaged groups
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
mobility can thus foster market-justice orientations, helping explain
how inequalities in access to welfare services become normatively
sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing accounts identify several mechanisms through which
intergenerational mobility may shape distributive justice preferences
but converge on two channels especially relevant for market-based
welfare: changes in material interests and shifts in normative frames
(\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Mobility alters individuals' material stakes by
changing their occupational status, income, and risk exposure. At the
same time, it reshapes their fairness judgments. Empirically, upward
mobility is often associated with stronger meritocratic interpretations
of success and fairness, whereas downward mobility is more closely
linked to structural explanations of inequality without necessarily
eroding meritocratic convictions (\citeproc{ref-bucca_merit_2016}{Bucca,
2016}; \citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). Regarding
market-based preferences, recent research shows that meritocratic
beliefs -- the conviction that effort and ability are rewarded -- are
strongly associated with support for market allocation
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). In such contexts, the upwardly mobile are especially likely to
view markets as just distributors, since their own trajectories appear
to confirm that markets appropriately reward effort; those who
experience downward mobility may still interpret outcomes through a
meritocratic lens, often reading losses as personal failure rather than
as evidence against market fairness. Thus, mobility and meritocracy
evolve on partly separate tracks yet intersect systematically
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Against this backdrop, the central aim of this article is to assess
whether intergenerational occupational mobility causally affects support
for the commodification of pensions. A second, more exploratory aim is
to examine whether this effect varies by meritocratic beliefs, using
this heterogeneity as an indication of possible underlying mechanisms. I
argue that social mobility operates through two main pathways
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Upward mobility
increases earnings, expected returns in a funded system, and perceived
economic security, making contribution-based access more appealing.
Simultaneously, upward moves reinforce individualistic, effort-based
understandings of success and, thus, the view that markets are just
assignors. Downward mobility tends to heighten economic insecurity,
dependence on public protection, and attachment to external attributions
of poverty, reducing support for market allocation. Meritocratic beliefs
are treated not as mediators but as moderators of these effects: among
strong meritocrats, upward mobility should amplify pro-commodification
attitudes, whereas downward mobility should attenuate -- or even reverse
-- its expected de-commodifying impact, as meritocratic interpretations
of effort lead individuals to view market outcomes as fair regardless of
direction. This inquiry contributes to research on market justice by
identifying the causal effects of mobility and using variation in
meritocratic beliefs to assess whether these effects follow a
meritocratic pathway.

This study focuses on Chile, an instructive case for examining
preferences for pension commodification. Despite sustained economic
growth, Chile combines high and persistent inequality with short-range
upward mobility and substantial barriers to higher-class positions
(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide in 1981 to fully replace a public pay-as-you-go
system with a mandatory, privately administered defined-contribution
scheme, while the state has progressively expanded a segmented
solidarity pillar to compensate for insufficient benefits
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 against the private pension
system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

@@

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  How does intergenerational social mobility affect individuals'
  preferences for the commodification of pensions in Chile?
\item
  How do meritocratic beliefs condition or moderate this relationship?
\end{enumerate}

To address these questions, the study draws on large-scale,
representative survey data for the urban Chilean population collected in
2016, 2018 and 2023. It adopts a causal framework of mobility effects
inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
The next section outlines the theoretical framework linking preferences
for market-based welfare, social mobility and expected heterogeneity by
meritocratic beliefs and time, and derives hypotheses. A subsequent
section presents the causal identification strategy, the data and the
analytical approach. The empirical sections report the findings, and the
article concludes by discussing what these results reveal about the
politics of market-based pension welfare and social mobility.

@@

Beyond the state's capacity to redistribute resources, market
institutions are central arenas for allocating socially valuable goods
and risks (\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Markets are not mere aggregates of individual choices but social
institutions endowed with rules and normative meanings, built through
``moral projects, saturated with normativity'' that permeate everyday
thinking around fairness, effort and deservingness
(\citeproc{ref-fourcade_moral_2007}{Fourcade \& Healy, 2007};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: shared norms and
beliefs about what is considered a just distribution, embedded and
reinforced through institutions and policy feedback
(\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). In an
era of privatisation and commodification, market logic has expanded into
core areas of social reproduction---childcare, healthcare, education and
pensions---stratifying access and quality
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}) and generating
feedback effects on public beliefs about welfare and market inequality
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In this context,
support for market-based welfare provision has grown worldwide,
particularly among higher-income groups who view private alternatives as
more efficient or higher-quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These institutional
transformations provide the backdrop for analysing how far citizens
accept the commodification of social protection
(\citeproc{ref-satz_por_2019}{Satz, 2019}).

The legitimacy of market-based welfare is closely tied to beliefs about
distributive justice grounded in market principles. Beyond
stratification structures, citizens hold structured beliefs about how
resources ought to be allocated, reflecting causal attributions for
inequality, normative principles and expectations about deservingness
(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}). Research
on distributive justice examines conceptions of how goods and rewards
should be distributed and when inequality is considered just
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}), while work on
redistributive preferences focuses on support for state-led mechanisms
that reduce inequality (\citeproc{ref-cavaille_fair_2025}{Cavaillé,
2025}). Situated at this intersection, the emerging literature on market
justice preferences conceptualises the market itself as a site of
redistribution and asks whether individuals regard market criteria as
fair bases for allocating wages and access to core welfare services
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In this
perspective, the expansion of market logic into welfare domains implies
that social services are treated as legitimate commodities that can be
priced and stratified (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}), and empirical work suggests that such institutional
changes feed back into attitudes, reinforcing market-conforming
understandings of fairness and responsibility
(\citeproc{ref-zhu_policy_2015}{Ling Zhu \& Lipsmeyer, 2015}).

Market justice preferences refer to normative beliefs that legitimise
allocating goods and services according to market criteria such as
income and purchasing power
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Building on Lane's
(\citeproc{ref-lane_market_1986}{1986}) contrast between market and
political justice, market justice is a distributive principle that
endorses rewarding effort, productivity and skill rather than need and
equality as emphasised in welfare-state policies. In this study, I treat
market justice less as an abstract doctrine and more concretely as
preferences for commodification: support for treating welfare services
as commodities, so that better services follow from higher income or
ability to pay, and the resulting stratified access is seen as fair
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Empirically, such
orientations are typically measured with items asking whether it is fair
that access to services depends on income, a strategy rooted in research
on justifications of capitalist inequality
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999}) and
applied to healthcare and education
(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}). Moving beyond single domains, Lindh
(\citeproc{ref-lindh_public_2015}{2015}) develops a comparative index of
income-based access to healthcare and education, while Castillo,
Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) use the
same single-item measure but add the pensions domain in Chile. These
constructs gauge how individuals view market-generated inequalities as
legitimate and capture two core dimensions of market distribution: the
role of income in determining attainment and the framing of services as
commodities that can be bought and sold according to ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

This study examines support for the commodification of pensions, defined
as normative judgements on whether better pensions should depend on
income or individual contributions. I treat this as a domain-specific
form of market justice: pensions are conceived as commodities whose
allocation should follow contribution- and return-based criteria rather
than principles of need (\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). The spread of funded,
privately managed schemes built on individual accounts has transformed
both the distributive and normative structure of old-age security
(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
Distributively, these schemes reassign risk from the collective to the
individual by tying benefits to formal employment, contribution density,
earnings and financial returns, making retirement income heavily
dependent on labour-market trajectories and capital-market success
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). Normatively, they recast
pensions as the outcome of personal investment decisions rather than
collective obligations, promoting ideals of individual responsibility,
self-economisation and contribution-based desert
(\citeproc{ref-arza_pension_2008}{Arza, 2008};
\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Contributions
are increasingly framed as private assets rather than risk pooling, and
workers are expected to act as employee--entrepreneurs who choose
providers, monitor portfolios and ``make their money work'' over the
life course (\citeproc{ref-borzutzky_chiles_2016}{Borzutzky \& Hyde,
2016}; \citeproc{ref-kerner_pension_2020}{Kerner, 2020};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As more people tie their
security expectations to private investments and financial markets,
their material interests become more deeply rooted in the market, and
their imperatives gain stronger normative support
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). When citizens are asked
whether higher pensions should be allocated to those who earn and
contribute more, they reflect support for the market allocation of a
critical social service and the legitimacy of the resulting risks and
inequalities in old age (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).

>>>>

\setstretch{1.5}
This document was last modified at 2025-12-01 02:26:24 and it was last
rendered at 2025-12-01 02:26:24.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions embed market-based criteria of ``fair'' allocation and
shape how citizens understand deservingness and the balance between
state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for the allocation of essential services such as
healthcare, education, and pensions
(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they legitimize unequal outcomes as the
product of individual responsibility and promote the conception of basic
social services as commodities.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}). As many countries expanded mandatory or voluntary
individual-capitalisation schemes---most radically in Chile---pensions
became organised around funded contribution-based entitlements and
market performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes not only how pension institutions allocate old-age risk and tie
benefit levels to lifetime labour-market trajectories
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also transforms the
justice principles citizens associate with old-age security: they must
judge whether higher benefits should follow market-derived criteria or
solidaristic principles
(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Although we know
something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: people with higher socioeconomic status are more
likely to support market-based criteria than disadvantaged groups
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial returns in funded schemes correlate with stronger support for
market-oriented pension provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources, and normative environments that reshape fairness judgements
(\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Social mobility can thus foster market-justice
orientations, helping explain how inequalities in access to welfare
services become normatively sustained
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Two channels are particularly relevant to how intergenerational mobility
can shape preferences for market-based welfare: changes in material
interests and shifts in normative frames
(\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Mobility alters individuals' material stakes by
changing their occupational status, income, and risk exposure. At the
same time, it reshapes their fairness judgments through acculturation
and learning experiences about the norms and values of the destination
class. Empirically, upward mobility is often associated with stronger
meritocratic interpretations of success and fairness, whereas downward
mobility is more closely linked to structural explanations of inequality
without necessarily eroding meritocratic convictions
(\citeproc{ref-bucca_merit_2016}{Bucca, 2016};
\citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). Regarding
market-based preferences, recent research shows that meritocratic
perceptions -- the conviction that effort is rewarded -- are strongly
associated with support for market allocation
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). In such contexts, it seems plausible that social mobility shapes
preferences for pension commodification through changes in their
material interests and in their meritocratic interpretations of
mobility. Thus, mobility and meritocracy evolve on partly separate
tracks yet intersect systematically
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Against this backdrop, the central aim of this article is to assess
whether intergenerational occupational mobility causally affects support
for the commodification of pensions. A second, more exploratory aim is
to examine whether this effect varies by meritocratic beliefs, using
this heterogeneity as an indication of possible underlying mechanisms. I
argue that social mobility operates through two main pathways
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Upward mobility
increases earnings, expected returns in a funded system, and perceived
economic security, making contribution-based access more appealing.
Simultaneously, upward moves reinforce individualistic, effort-based
understandings of success and, thus, the view that markets are just
assignors. Downward mobility tends to heighten economic insecurity,
dependence on public protection, and attachment to external attributions
of poverty, reducing support for market allocation. Meritocratic beliefs
are treated not as mediators but as moderators of these effects: among
strong meritocrats, upward mobility should amplify pro-commodification
attitudes, whereas downward mobility should attenuate -- or even reverse
-- its expected de-commodifying impact, as meritocratic interpretations
of effort lead individuals to view market outcomes as fair regardless of
direction. This inquiry contributes to research on market justice by
identifying the causal effects of social mobility and the role of
meritocratic beliefs.

This study focuses on Chile, an instructive case for examining
preferences for pension commodification. Despite sustained economic
growth, Chile combines high and persistent inequality with short-range
upward mobility and substantial barriers to higher-class positions
(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide in 1981 to fully replace a public pay-as-you-go
system with a mandatory, privately administered defined-contribution
scheme, while the state has progressively expanded a segmented
solidarity pillar to compensate for insufficient benefits
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2016 to 2019 against the private pension
system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security, and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

@@

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  How does intergenerational social mobility affect individuals'
  preferences for the commodification of pensions in Chile?
\item
  How do meritocratic beliefs condition this relationship?
\end{enumerate}

To address these questions, the study draws on large-scale,
representative survey data for the urban Chilean population collected in
2016, 2018, and 2023. It adopts a causal framework of mobility effects
inspired by Breen \& Ermisch (\citeproc{ref-breen_effects_2024}{2024}).
The following section outlines the theoretical framework linking
preferences for pension welfare commodification, social mobility, and
expected heterogeneity by meritocratic beliefs and derives hypotheses. A
subsequent section presents the causal identification strategy, the
data, and the analytical approach. The empirical sections report the
findings, and the article concludes by discussing what these results
reveal about the politics of market-based pension welfare and social
mobility.

@@

Beyond the state's capacity to redistribute resources, market
institutions are central arenas for allocating socially valuable goods
and risks (\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Markets are not mere aggregates of individual choices but social
institutions endowed with rules and normative meanings, built through
``moral projects, saturated with normativity'' that permeate everyday
thinking around fairness, effort and deservingness
(\citeproc{ref-fourcade_moral_2007}{Fourcade \& Healy, 2007};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: shared norms and
beliefs about what is considered a just distribution, embedded and
reinforced through institutions and policy feedback
(\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). In an
era of privatisation and commodification, market logic has expanded into
core areas of social reproduction---childcare, healthcare, education and
pensions---stratifying access and quality
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}) and generating
feedback effects on public beliefs about welfare and market inequality
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In this context,
support for market-based welfare services provision has grown worldwide,
particularly among higher-income groups who view private alternatives as
more efficient or higher-quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These institutional
transformations provide the backdrop for analysing how far citizens
accept the commodification of social protection
(\citeproc{ref-satz_por_2019}{Satz, 2019}).

The legitimacy of market-based welfare is closely tied to beliefs about
distributive justice grounded in market principles. Beyond
stratification structures, citizens hold structured beliefs about how
resources ought to be allocated, reflecting causal attributions for
inequality, normative principles and expectations about deservingness
(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}). Research
on distributive justice examines conceptions of how goods and rewards
should be distributed and when inequality is considered just
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}), while work on
redistributive preferences focuses on support for state-led mechanisms
that reduce inequality (\citeproc{ref-cavaille_fair_2025}{Cavaillé,
2025}). Situated at this intersection, the emerging literature on market
justice preferences conceptualises the market itself as a site of
redistribution and asks whether individuals regard market criteria as
fair bases for allocating wages and access to core welfare services
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In this
perspective, the expansion of market logic into welfare domains implies
that social services are treated as legitimate commodities that can be
priced and stratified (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}), and empirical work suggests that such institutional
changes feed back into attitudes, reinforcing market-conforming
understandings of fairness and welfare
(\citeproc{ref-zhu_policy_2015}{Ling Zhu \& Lipsmeyer, 2015}).

Market justice preferences refer to normative beliefs that legitimise
allocating goods and services according to market criteria such as
income and purchasing power
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Building on Lane's
(\citeproc{ref-lane_market_1986}{1986}) contrast between market and
political justice, market justice is a distributive principle that
endorses rewarding effort, productivity and skill rather than need and
equality as emphasised in welfare-state policies. In this study, I treat
market justice less as an abstract doctrine and more concretely as
preferences for commodification: support for treating welfare services
as commodities, so that better services follow from higher income or
ability to pay, and the resulting stratified access is seen as fair
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Empirically, such
orientations are typically measured with survey items asking whether it
is fair that access to better services depends on income, a strategy
rooted in research on justifications of capitalist inequality
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999}) and
applied to healthcare and education
(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}). Moving beyond single domains, Lindh
(\citeproc{ref-lindh_public_2015}{2015}) develops a comparative index of
income-based access to healthcare and education, while Castillo,
Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) use the
same single-item measure but add the pensions domain in Chile. These
constructs gauge how individuals view market-generated inequalities as
legitimate and capture two core dimensions of market distribution: the
role of income in determining attainment and the framing of services as
commodities that can be bought and sold according to ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

This study examines support for the commodification of pensions, defined
as normative judgements on whether better pensions should depend on
income or individual contributions. I treat this as a domain-specific
form of market justice: pensions are conceived as commodities whose
allocation should follow contribution- and return-based, and whose
outcome is seen as fair (\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). The spread of funded,
privately managed schemes built on individual capitalisation accounts
has transformed both the distributive and normative structure of old-age
security (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
Distributively, these schemes reassign risk from the collective to the
individual by tying benefits to formal employment, contribution density,
earnings and financial returns, making retirement income heavily
dependent on labour-market trajectories and capital-market success
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). Normatively, they recast
pensions as the outcome of personal investment decisions rather than
collective obligations, promoting ideals of individual responsibility,
self-economisation and contribution-based desert
(\citeproc{ref-arza_pension_2008}{Arza, 2008};
\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Contributions
are increasingly framed as private assets rather than risk pooling, and
workers are expected to act as employee--entrepreneurs who choose
providers, monitor portfolios and ``make their money work'' over the
life course (\citeproc{ref-borzutzky_chiles_2016}{Borzutzky \& Hyde,
2016}; \citeproc{ref-kerner_pension_2020}{Kerner, 2020};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As more people tie their
security expectations to private investments and financial markets,
their material interests become more deeply rooted in the market, and
their imperatives gain stronger normative support
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). When citizens are asked
whether higher pensions should be allocated to those who earn and
contribute more, they reflect support for the market allocation of a
critical social service and the legitimacy of the resulting risks and
inequalities in old age (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).
